A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-14 tex/codatadp.tex
edits here and there in Sects. 1 and 2

\begin{abstract}
Codatatypes naturally capture potentially infinite data structures and
processes. We present a decision procedure that combines reasoning about
datatypes and codatatypes. The dual of the cyclicity rule for datatypes is a
bisimilarity rule that identifies equal values, also in the presence of cycles.
The procedure is complete for universal problems and is compatible with the
Nelson--Oppen method. It has been implemented in the latest version of CVC4, a
state-of-the-art SMT solver. An evaluation based on problems generated from
Isabelle theories demonstrates the potential of the procedure.
\end{abstract}

@@

Freely generated inductive datatypes are ubiquitous in functional programs and
logical specifications. They are especially
useful to represent finite data structures in computer science applications but
also arise in mathematical formalizations.
They can be implemented efficiently and enjoy
properties that can be exploited in automated reasoners. 
%
%However, because datatype values correspond to finite ground terms, they
%are generally not adequate to represent infinite objects.
%For example, the datatype of natural numbers
%constructed by $\const{Z} : \ty{nat}$ and $\const{S} : \ty{nat} \to \ty{nat}$
%only allow values of the form $\const{S}(\ldots(\const{S}(\const{Z}))\ldots)$.

To represent infinite values, % such as $\const{S}(\const{S}(\const{S}(\ldots))$,
a natural choice is to turn to \emph{codatatypes} (or coinductive datatypes), the
non-well-founded dual of (inductive) \relax{datatypes}.
%
Although they are sometimes perceived as esoteric, codatatypes have an important
role to play in computer science. The verified C compiler CompCert
\cite{leroy-2009}, the verified Java compiler Jinja\-Threads
\cite{lochbihler-2010-jinja}, and the formalized Java memory model
\cite{lochbihler-2014-jmm} all depend on codatatypes to capture infinite
processes.
% as do the CoALP dialect of Prolog and the Dafny specification language.

Codatatypes are freely generated by their constructors, but in contrast with datatypes,
infinit\-e constructor terms are also legitimate values for codatatypes
(Section~\ref{sec:the-theory-of-co-datatypes}). Intuitively, the
values of a codatatype consist of all well-typed finite and infinite ground constructor
terms, and only those. For example, the specification
%
\[\keyw{codatatype}~\,\ty{enat} \,=\, \const{Z} \,\mid\, \const{S}(\ty{enat})\]
%
(using an ML-like syntax) introduces a type that
models the natural numbers $\const{Z}$, $\const{S}(\const{Z})$, $\const{S}(\const{S}(\const{Z}))$, $\ldots$\afterLdots{},
using Peano notation but extended with an
infinity value $\infty = \const{S}(\const{S}(\const{S}(\ldots)))$. Compared
with the more conventional, inductive definition
\[\keyw{datatype}~\,\ty{enat} \,=\, \const{Z} \,\mid\, \const{S}(\ty{enat}) \,\mid\, \const{Infty}\]
%
the codatatype avoids one case by unifying the finite and infinite nonzero cases.
Moreover, equations such as $\const{S}(\infty) \teq \infty$ hold without additional
work, because both sides expand to the infinite term
$\const{S}(\const{S}(\const{S}(\ldots)))$, which uniquely identifies the
value $\infty$.

@@

Datatypes and codatatypes are an integral part of most popular proof assistants,
including Agda, Coq, Isabelle, Matita, and PVS. In recent years, some
automatic theorem provers have been extended with dedicated reasoning facilities
for datatypes. The SMT-LIB~2 \cite{barrett-et-al-2010} syntax, implemented by
most SMT (satisfiability modulo theories) solvers, includes a theory of
datatypes.

In this \thewordpaper, we introduce a unified decision procedure for ground
problems involving datatypes and codatatypes in combination and discuss extensions 
of the procedure for handling universal (non-ground) problems.
The procedure is
described abstractly as a calculus (Section~\ref{sec:the-calculus}) and is
designed to be composable via the Nelson--Oppen method \cite{nelson-oppen-1979}.
For the datatype case, it follows the lines of Barrett et al.\ \cite{barrett-et-al-2007}.
To our knowledge, our procedure is the first of its kind for the theory of
codatatypes.

@@

Third, \emph{it must be possible to express cyclic }(\emph{$\omega$-regular}) \emph{values as closed terms and
to enumerate them.} This is necessary both for finite model finding (modulo theories)
and for theory combinations. The $\mu$-binder notation is used to represent cyclic
values internally. For example,
the term $\const{SCons}(1,\: \mu\vthinspace s.\; \const{SCons}(0,\: \const{SCons}(9,\: s)))$
denotes the lasso-shaped sequence $1, 0, 9, 0, 9, 0, 9, \ldots$\afterLdots.

@@

\begin{paper}
Detailed proofs of soundness and completeness are included in the technical report
associated with this paper \cite{our-report}.
\end{paper}%
The procedure is implemented in the SMT solver CVC4 as a combination
of rewriting and a theory solver
(Section~\ref{sec:the-theory-solver}).
It consists of about 2000 lines of \cpp{} code, among which 1600 are
shared between datatypes and codatatypes. The code is integrated in the
development version of the solver and is expected to be part of the CVC4~1.5 release.
%
An evaluation on %hand-crafted examples and on
problems generated from Isabelle theories using the Sledgehammer tool
demonstrates the usefulness of the approach (Section~\ref{sec:experimental-results}).

@@

%  * setting: FOL
%    * core procedure is restricted to ground
%      * but theory solver cooperates
%    * still, universal conjectures fall into the ground fragment via negation
%      and skolemization (or, alternatively, herbrandization)

@@

\paragraph{Related Work.}
Although it was written several years ago, the account of related work in
Barrett et al.\ \cite{barrett-et-al-2007} is still a good reference.
Since then, datatypes have been added to Z3 in unpublished work by Leonardo de
Moura and to a SPASS-like prototype prover called Pirate by Daniel Wand.
%\cite{wand-weidenbach-201x}.
Closely related is the automatic structural
induction in CVC4 by Reynolds and Kuncak \cite{reynolds-kuncak-2015} and in
Pirate by Wand and Weidenbach \cite{wand-weidenbach-201x}, which depends on a
notion of datatype.

@@

Function symbols are written in a sans-serif font (e.g., $\const{f}$, $\const{g}$) to
distinguish them from variables (e.g., $x$, $y$). Symbols starting with an uppercase
letter (e.g.,
$\const{S}$, $\const{SCons}$) are reserved for constructors. With each function symbol \const{f}
is associated a list of argument types $\tau_1,\ldots,\tau_n$ (with $n \ge 0$)
and a return type $\tau$. This connection can be expressed compactly as
$\const{f} : \tau_1 \times \cdots \times \tau_n \to \tau$,
which collapses to $\const{f} : \tau$ if $n = 0$.
For a term $t$, the notation $t : \tau$ indicates that it has type $\tau$.
Functions are invoked in the standard way, with $\const{f}(t_1,\ldots,t_n)$
applying the $n$-ary function symbol
\const{f} to $n$ well-typed arguments $t_1 :\nobreak \tau_1$, \ldots, $t_n :
\tau_n$. Nullary function symbols are called constants and can appear without
parentheses in terms.
%
Finally, the notation $\bar x$ abbreviates a list or tuple $x_1,\ldots,x_n$
and $\bigwedge_{\,i}\, \varphi_i$ abbeviates a conjunction
$\varphi_1 \mathrel\land \cdots \mathrel\land \varphi_n$.

@@

We fix a signature $\Sig = (\Types, \Funcs)$. The types are partioned into
$\Types = \Data \mathrel{\uplus} \Codata \mathrel{\uplus} \Nondata$, where $\Data$ are the
\emph{datatypes}, $\Codata$ are the \emph{codatatypes}, and $\Nondata$ are the remaining
\emph{ordinary types}. The functions are partitioned into $\Funcs = \Ctr
\mathrel{\uplus} \Sel$, where $\Ctr$ are the \emph{constructors} and $\Sel$ are the
\emph{selectors}; there is no need to consider further function symbols
because they can be abstracted away as variables when combining theories.

@@

Because the types are inhabited (nonempty), a datatype specification is
admissible only if a ground constructor term can be exhibited that belongs to
the datatype. This rules out non-well-founded specifications such as
\[\keyw{datatype}\;\, \ty{fstream}_{\,\tau} \,=\, \const{FSCons}(\tau,\:\, \ty{fstream}_{\,\tau})\]
For codatatypes, no admissibility check is necessary because there is always a term,
finite or infinite, that witnesses nonemptiness.
%Given the specification
%
%\[\keyw{codatatype}\;\, \ty{stream} \,=\, \const{SCons}(\ty{int},\:\ty{stream})\]
%
%the infinite value $\const{SCons}(0, \const{SCons}(0, \ldots))$ witnesses
%nonemptiness.

@@

Datatypes are additionally characterized by an induction axiom schema:
%
\[
\begin{aligned}[t]
\text{Induction:}\quad
\AXC{\strut$\bigwedge_{\,i,j}\, \forall x_1 \ldots \vthinspace x_{n_{i\negvthinspace j}}.\; \bigl(\bigwedge_{\,k}\, \mathit{IH}_{i\negvthinspace j}^k[x_k]\bigr) \longrightarrow P_i[C_{i\negvthinspace j}(x_1,\ldots,x_{n_{ij}})]$}
\UIC{\strut$\bigwedge_{\,i}\, P_i[v_i]$}
\DP
\end{aligned}
\]
where the induction hypothesis $\mathit{IH}_{i\negvthinspace j}^k(x)$
denotes either $P_{\iPrime}(x)$ if there exists some $\iPrime$ such that
the formula is type-correct or else $\top$.
%
The axiom schema ensures that the interpretation of datatypes
is a standard model. For example,
for a datatype of natural numbers constructed from $\const{Z}$ and $\const{S}$,
induction rules out nonstandard models, which could contain cyclic values---e.g.,
an $n$ such that $n \teq \const{S}(n)$---or even infinite acyclic values
$\const{S}(\const{S}(\ldots))$.

@@

It is worth observing that nothing forbids non(co)recursive specifications
such~as
\[\begin{aligned}[t]
      \keyw{datatype}\;\, \ty{ext\_nat} & \,=\, \const{Nat}(\ty{nat}) \,\mid\, \const{Infty} \\[-.5\jot]
      \keyw{codatatype}\; \,\ty{complex} & \,=\, \const{Complex}(\const{re}{:}\; \ty{real},\; \const{im}{:}\;\ty{real})
\end{aligned}
\]
At the semantic level, it makes no difference whether the types are introduced as
datatypes or as codatatypes.
%Without loss of generality, we consider that these types are datatypes
%instead of codatatypes.

Finally, some codatatypes are so degenerate as to have infinite values
even though they are finite. The simplest example is
\[\keyw{codatatype}\;\, \ty{a} \,=\, \const{A}(\ty{a})\]
This is a finite codatatype, of cardinality one, whose single value is
infinite: \,$\mu\vthinspace a.\; A(a)$. Other specimens are
\[\begin{aligned}[t]
      \keyw{codatatype}\;\, \ty{b} & \,=\, \const{B}(\ty{b},\: \ty{c},\: \ty{b},\: \ty{unit}) \\[-\jot]
             \keyw{and}\;\, \ty{c} & \,=\, \const{C}(\ty{a},\: \ty{unit},\: \ty{b},\: \ty{c})
\end{aligned}
\]
where \ty{unit} is a datatype with the single constructor $\const{Unity} :
\ty{unit}$, as well as $\ty{stream}_{\,\ty{unit}\,}$. We call such types
\emph{corecursive singletons}. For the decision procedure, it will be
important to detect these types, even if they occur rarely in practice. A type
may also be a corecursive singleton only in some models. For example, if
\ty{unit} is an uninterpreted type, \ty{b} and \ty{c} will be singleton types
precisely in the models that interpret \ty{unit} as a singleton. Fortunately,
given some cardinalities for the ordinary types, it is easy to characterize
this degenerate case:

@@

\begin{lemma}[Corecursive Singletons]%
\label{lem:corecursive-singletons}%
\afterDot
Let $\delta$ be a corecursive codatatype. The domain interpreting $\delta$ is
either infinite or a singleton. In the latter case, $\delta$ necessarily has a
single constructor, whose arguments have types that are interpreted by
singletons.
\end{lemma}

@@

Therefore, $\const{C}$ must be the only constructor.
If any of its noncorecursive arguments has a cardinality greater than 1,
it is possible to encode alternation patterns using it---e.g.,
$\const{C}(0,\: \const{C}(1,\: \const{C}(0,\: \const{C}(0,\: \ldots))))$---which
again excludes finite models. Otherwise, the coinduction principle ensures
that the type has only one value.
\qed
\end{proof}

@@

\section{A Ground Decision Procedure for (Co)-Datatypes}
\label{sec:the-calculus}

@@

\section{Extension to Quantified Formulas}

>>>>

\begin{abstract}
Codatatypes naturally capture potentially infinite data structures and
processes. We present a decision procedure that combines reasoning about
datatypes and codatatypes. The dual of the cyclicity rule for datatypes is a
bisimilarity rule that identifies equal values, also in the presence of cyclic
($\omega$-regular) data. The procedure is complete for ground problems and is
composable via the Nelson--Oppen method. It has been implemented in the latest
version of CVC4, a state-of-the-art SMT solver. An evaluation based on
problems generated from Isabelle theories demonstrates the potential of the
procedure.
\end{abstract}

@@

Freely generated inductive datatypes are ubiquitous in functional programs and
logical specifications. They are especially
useful to represent finite data structures in computer science applications but
also arise when formalizing mathematics.
They can be implemented efficiently and enjoy
properties that can be exploited in automated reasoners. 
%
%However, because datatype values correspond to finite ground terms, they
%are generally not adequate to represent infinite objects.
%For example, the datatype of natural numbers
%constructed by $\const{Z} : \ty{nat}$ and $\const{S} : \ty{nat} \to \ty{nat}$
%only allow values of the form $\const{S}(\ldots(\const{S}(\const{Z}))\ldots)$.

To represent infinite values, % such as $\const{S}(\const{S}(\const{S}(\ldots))$,
a natural choice is to turn to coinductive datatypes, or \emph{codatatypes},
the non-well-founded dual of inductive \emph{datatypes}.
%
Although they are sometimes perceived as esoteric, codatatypes have an important
role to play in computer science. The verified C compiler CompCert
\cite{leroy-2009}, the verified Java compiler Jinja\-Threads
\cite{lochbihler-2010-jinja}, and the formalized Java memory model
\cite{lochbihler-2014-jmm} all depend on codatatypes to capture infinite
processes.

Codatatypes are freely generated by their constructors, but in contrast with datatypes,
infinit\-e constructor terms are also legitimate values for codatatypes
(Section~\ref{sec:the-theory-of-co-datatypes}). Intuitively, the
values of a codatatype consist of all well-typed finite and infinite ground constructor
terms, and only those. For example, the coinductive specification
%
\[\keyw{codatatype}~\,\ty{enat} \,=\, \const{Z} \,\mid\, \const{S}(\ty{enat})\]
%
(using an ML-like syntax) introduces a type that
models the natural numbers $\const{Z}$, $\const{S}(\const{Z})$, $\const{S}(\const{S}(\const{Z}))$, $\ldots$\afterLdots{},
using Peano notation but extended with an
infinity value $\infty = \const{S}(\const{S}(\const{S}(\ldots)))$. Compared
with the more conventional, inductive definition
\[\keyw{datatype}~\,\ty{enat} \,=\, \const{Z} \,\mid\, \const{S}(\ty{enat}) \,\mid\, \const{Infty}\]
%
the codatatype avoids one case by unifying the finite and infinite nonzero cases.
Moreover, equations such as $\const{S}(\infty) \teq \infty$ hold without additional
work, because both sides expand to the infinite term
$\const{S}(\const{S}(\const{S}(\ldots)))$, which uniquely identifies the
value $\infty$.

@@

Datatypes and codatatypes are an integral part of modern proof assistants,
including Agda, Coq, Isabelle, Matita, and PVS. In recent years, some
automatic theorem provers have been extended with dedicated reasoning facilities
for datatypes. The SMT-LIB~2 \cite{barrett-et-al-2010} syntax, implemented by
most SMT (satisfiability modulo theories) solvers, includes a theory of
datatypes.

In this \thewordpaper, we introduce a unified decision procedure for ground
problems involving datatypes and codatatypes in combination
(Section~\ref{sec:a-ground-decision-procedure-for-co-datatypes})
and discuss extensions  of the procedure for handling quantifiers
(Section~\ref{sec:extension-to-quantified-formulas}).
The procedure is described abstractly as a calculus and is
designed to be composable via the Nelson--Oppen method \cite{nelson-oppen-1979}.
For the datatype case, it follows the lines of Barrett et al.\ \cite{barrett-et-al-2007}.
To our knowledge, our procedure is the first of its kind for the theory of
codatatypes.

@@

Third, \emph{it must be possible to express cyclic }(\emph{$\omega$-regular}) \emph{values as closed terms and
to enumerate them.} This is necessary both for finite model finding (modulo theories)
and for theory combinations. The $\mu$-binder notation is used to represent cyclic
values internally. For example,
the term $\const{SCons}(1,\: \mu s.\; \const{SCons}(0,\: \const{SCons}(9,\: s)))$
denotes the lasso-shaped sequence $1, 0, 9, 0, 9, 0, 9, \ldots$\afterLdots.

@@

\begin{paper}
Detailed proofs of soundness and completeness are included in the technical report
associated with this paper \cite{our-report}.
\end{paper}%
The procedure is implemented in the SMT solver CVC4 as a combination
of rewriting and a theory solver
(Section~\ref{sec:the-theory-solver}).
\textbf{TODO: CHECK WITH IMPLEMENTATION, but round to nearest 50 or 1000.}
It consists of about 2000 lines of \cpp{} code, among which 1600 are
shared between datatypes and codatatypes. The code is integrated in the
development version of the solver and is expected to be part of the CVC4~1.5 release.
%
An evaluation on %hand-crafted examples and on
problems generated from Isabelle theories using the Sledgehammer tool
demonstrates the usefulness of the approach (Section~\ref{sec:experimental-results}).

@@

%  * setting: FOL
%    * core procedure is restricted to ground
%      * but theory solver cooperates

@@

\paragraph{Related Work.}
Although it was written several years ago, the account of related work in
Barrett et al.\ \cite{barrett-et-al-2007} is still a good starting point.
Since then, datatypes have been added to the SMT solver
Z3 in unpublished work by Leonardo de Moura and to a SPASS-like prototype
superposition prover called Pirate by Daniel Wand.
Closely related is the automatic structural
induction in CVC4 by Reynolds and Kuncak \cite{reynolds-kuncak-2015} and in
Pirate by Wand and Weidenbach \cite{wand-weidenbach-201x};
both methods naturally depend on a notion of datatype.

@@

Function symbols are written in a sans-serif font (e.g., $\const{f}$, $\const{g}$) to
distinguish them from variables (e.g., $x$, $y$). Symbols starting with an uppercase
letter (e.g.,
$\const{S}$, $\const{SCons}$) are reserved for constructors. With each function symbol \const{f}
is associated a list of argument types $\tau_1,\ldots,\tau_n$ (with $n \ge 0$)
and a return type $\tau$. This connection can be expressed %compactly
as $\const{f} : \tau_1 \times \cdots \times \tau_n \to \tau$,
which collapses to $\const{f} : \tau$ if $n = 0$.
For a term $t$, the notation $t : \tau$ indicates that it has type $\tau$.
Functions are invoked in the standard way, with $\const{f}(t_1,\ldots,t_n)$
applying the $n$-ary function symbol
\const{f} to $n$ well-typed arguments $t_1 :\nobreak \tau_1$, \ldots, $t_n :
\tau_n$. Nullary function symbols are called constants and can appear without
parentheses in terms.
%
Finally, the notation $\bar x$ abbreviates a list or tuple $x_1,\ldots,x_n$
and $\bigwedge_{\,i}\, \varphi_i$ abbeviates a conjunction
$\varphi_1 \mathrel\land \cdots \mathrel\land \varphi_n$.

@@

We fix a signature $\Sig = (\Types, \Funcs)$. The types are partioned into
$\Types = \Data \mathrel{\uplus} \Codata \mathrel{\uplus} \Nondata$, where $\Data$ are the
\emph{datatypes}, $\Codata$ are the \emph{codatatypes}, and $\Nondata$ are the %remaining
\emph{ordinary types}. The functions are partitioned into $\Funcs = \Ctr
\mathrel{\uplus} \Sel$, where $\Ctr$ are the \emph{constructors} and $\Sel$ are the
\emph{selectors}; there is no need to consider further function symbols
because they can be abstracted away as variables when combining theories.

@@

Because all types must be inhabited (nonempty), a datatype specification is
admissible only if a ground constructor term can be exhibited that belongs to
the datatype. This rules out non-well-founded specifications such as
\[\keyw{datatype}\;\, \ty{fstream}_{\,\tau} \,=\, \const{FSCons}(\tau,\:\, \ty{fstream}_{\,\tau})\]
For codatatypes, no admissibility check is necessary because there is always a term,
finite or infinite, that witnesses nonemptiness.
%Given the specification
%
%\[\keyw{codatatype}\;\, \ty{stream} \,=\, \const{SCons}(\ty{int},\:\ty{stream})\]
%
%the infinite value $\const{SCons}(0, \const{SCons}(0, \ldots))$ witnesses
%nonemptiness.

@@

Datatypes are additionally characterized by an induction axiom schema:
%
\[
\begin{aligned}[t]
\text{Induction:}\quad
\AXC{\strut$\bigwedge_{\,i,j}\, \forall x_1 \ldots \vthinspace x_{n_{i\negvthinspace j}}.\; \bigl(\bigwedge_{\,k}\, \mathit{IH}_{i\negvthinspace j}^k[x_k]\bigr) \longrightarrow P_i[C_{i\negvthinspace j}(x_1,\ldots,x_{n_{ij}})]$}
\UIC{\strut$\bigwedge_{\,i}\, P_i[v_i]$}
\DP
\end{aligned}
\]
where the induction hypothesis $\mathit{IH}_{i\negvthinspace j}^k(x)$
denotes either $P_{\iPrime}(x)$ if there exists some $\iPrime$ such that
the formula is type-correct or else $\top$.
%
The axiom schema ensures that the interpretation of datatypes
is a standard model. For example,
for a datatype of natural numbers constructed from $\const{Z}$ and $\const{S}$,
induction prohibits nonstandard models, which could contain cyclic values---e.g.,
an $n$ such that $n \teq \const{S}(n)$---or even infinite acyclic values
$\const{S}(\const{S}(\ldots))$.

@@

It is worth observing that nothing forbids non(co)recursive specifications
such~as
\[\begin{aligned}[t]
      \keyw{datatype}\;\, \ty{enat} & \,=\, \const{Nat}(\ty{nat}) \,\mid\, \const{Infty} \\[-.5\jot]
      \keyw{codatatype}\; \,\ty{complex} & \,=\, \const{Complex}(\const{re}{:}\; \ty{real},\; \const{im}{:}\;\ty{real})
\end{aligned}
\]
At the semantic level, it makes no difference whether such types are
introduced as datatypes or as codatatypes.
%Without loss of generality, we consider that these types are datatypes
%instead of codatatypes.

Finally, some codatatypes are so degenerate as to have infinite values
even though they are finite. The simplest example is
\[\keyw{codatatype}\;\, \ty{a} \,=\, \const{A}(\ty{a})\]
This is a finite codatatype, of cardinality one, whose single value is
infinite: \,$\mu a.\; A(a)$. Other specimens are
\[\begin{aligned}[t]
      \keyw{codatatype}\;\, \ty{b} & \,=\, \const{B}(\ty{b},\: \ty{c},\: \ty{b},\: \ty{unit}) \\[-\jot]
             \keyw{and}\;\, \ty{c} & \,=\, \const{C}(\ty{a},\: \ty{unit},\: \ty{b},\: \ty{c})
\end{aligned}
\]
where \ty{unit} is a datatype with the single constructor $\const{Unity} :
\ty{unit}$, as well as $\ty{stream}_{\,\ty{unit}\,}$. We call such types
\emph{corecursive singletons}. For the decision procedure, it will be
important to detect these types, even if they occur rarely in practice. A type
may also be a corecursive singleton only in some models. For example, if
\ty{unit} is an uninterpreted type, \ty{b} and \ty{c} will be singleton types
precisely for the models that interpret \ty{unit} as a singleton. Fortunately,
given cardinalities for the ordinary types, it is easy to characterize
this degenerate case:

@@

\begin{lemma}[Corecursive Singletons]%
\label{lem:corecursive-singletons}%
\afterDot
Let $\delta$ be a corecursive codatatype. The domain interpreting $\delta$ is
either infinite or a singleton. In the latter case, $\delta$ necessarily has a
single constructor, whose arguments have types that are interpreted as
singletons.
\end{lemma}

@@

Therefore, $\const{C}$ must be the only constructor.
If any of its noncorecursive arguments has a cardinality greater than 1,
it is possible to encode alternation patterns using it---e.g.,
$\const{C}(0,\: \const{C}(1,\: \const{C}(0,\: \const{C}(0,\: \ldots))))$---which
again excludes finite models. Otherwise, the coinduction principle ensures
that the type has at most one value.
\qed
\end{proof}

@@

\section{A Ground Decision Procedure for (Co)datatypes}
\label{sec:a-ground-decision-procedure-for-co-datatypes}

@@

\section{Extension to Quantified Formulas}
\label{sec:extension-to-quantified-formulas}

  * universal conjecture is no problem:
    * falls into the ground fragment via negation
      and skolemization
  * nor is existential axiom a problem
  * what's interesting: universal axioms and existential conjectures
